---
title: "Comparison of the Computational Costs of Various Sampling and Prediction Methods for Gaussian Processes with Matérn Covariance Functions"
author: "David Bolin, Vaibhav Mehandiratta, and Alexandre B. Simas"
date: "Created: 2024-05-20. Last modified: `r Sys.Date()`."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparison of different methods for sampling and doing prediction for Gaussian processes with Matérn covariance function}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
references:
- id: BMS2024
  title: "Linear cost and exponentially convergent approximation of Gaussian Matérn stochastic processes"
  author:
  - family: Bolin
    given: David
  - family: Mehandiratta
    given: Vaibhav
  - family: Simas
    given: Alexandre B.
  container-title: Preprint
- id: DBFG
  title: "Hierarchical nearest-neighbor Gaussian process models for large geostatistical datasets"
  author:
  - family: Datta
    given: A.
  - family: Benerjee
    given: S. 
  - family: Finley
    given: A. O.
  - family: Gelfand
    given: A. E.
  container-title: J. Amer. Statist. Assoc.
  type: article
  issue: 111
  pages: 800-812
  issued:
    year: 2016
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(1)
```

## Introduction

This vignette contains some time comparisons for sampling and doing prediction of stationary Gaussian processes with the Matérn covariance function
$$r(h) = \frac{\sigma^2}{2^{\nu-1}\Gamma(\nu)}(\kappa h)^\nu K_\nu(\kappa h),$$
where $\Gamma(\cdot)$ is the Gamma function and $K_\nu$ is the modified Bessel function of the second kind, $\kappa$ is connected to the range of the process, $\nu$ is the smoothness parameter and $\sigma$ is the marginal standard deviation. We will consider some approximations, the reference approximation is the one based on the method introduced in [@BMS2024]. We will compare it against a singular value decomposition method (PCA) and a nearest neighbor approach [@DBFG]. We will use a calibration of the hyperparameter $m$ introduced in [@BMS2024] based on the theoretical costs obtained in Table 1 of [@BMS2024]. 

## Preprocessing

We start by installing the newest version of rSPDE to make sure all the functions we will are are available. 

```{r,message=FALSE}
devtools::install_github("davidbolin/rSPDE")
```

We will now load all the auxiliary functions:

```{r, message=FALSE}
source("aux_functions/aux_functions_cov.R")
source("aux_functions/predict_methods.R")
source("aux_functions/prediction_computations.R")
source("aux_functions/simulate_methods.R")
```

## Cost comparison based on predictions

For this case we will consider that we have $N=1000$ observations, the range parameter will be $0.2$, the standard deviation will be $1$. The cost will be measured for the smoothness parameter $\nu$ between $0$ and $0.5$, between $0.5$ and $1.5$, and between $1.5$ and $2.5$. The hyperparameter of the model will be $m$ taken from 1 to 6 for the reference method, and calibrated following Table 1 of [@BMS2024] for the PCA method and nearest neighbor methods. 

We start by defining the parameters:

```{r}
N <- c(1000,2000)
n_obs <- c(800, 1800)
# L - length of the interval, [0,L]
L <- 30
range <- 1
sigma <- 1
sigma_e <- 0.1
m_rat <- 1:6
m_nngp_fun <- function(m, alpha){
              if(alpha<1) {
                mn <- m
            } else if (alpha < 2) {
                mn <- round(m*(ceil(alpha)+1)^1.5)    
            } else {
                mn <- round(m*(ceil(alpha)+1)^1.5)
            }
            return(mn)
} 
m_pca_fun <- function(m, alpha){
      return(pmax(9*round((m+1)*sqrt(m) * ceil(alpha)^(3/2)),2))
}
```

We will start by comparing the times and errors for our method, NNGP and KL. We will observe in `n_obs` random locations and predict on an equally spaced grid. To this end we will use the auxiliary `compare_times()` function. We will obtain 10 samples.

```{r}
results <- list()
samples <- 10
for(sample in 1:samples){
  cat(sample, '\n')
  results[[sample]] <- compare_times(N, n_obs, L, range, sigma, sigma_e, m_rat, m_nngp_fun, m_pca_fun)
}
```



## Cost comparison based on sampling

We will now compare the times for the same choices of parameters by using the `compare_times_simulation()` function:

```{r}
m_nngp_fun_sim <- function(m, alpha){
      m_nngp <- round(8*(m+1)* ceil(alpha+1))
      d_m <- diff(m_nngp)
      d_m <- ifelse(d_m == 0, 1, d_m)
      m_nngp <- c(m_nngp[1], m_nngp[1]+cumsum(d_m))
      return(pmax(m_nngp,1))
} 
m_pca_fun_sim <- function(m, alpha){
        m_kl <- round(8*(m+1) * ceil(alpha+1))
        d_m <- diff(m_kl)
        d_m <- ifelse(d_m == 0, 1, d_m)
        m_kl <- c(m_kl[1], m_kl[1]+cumsum(d_m))
        m_kl <- m_kl + 1
        return(pmax(m_kl,1))
}
```

```{r, results="hide"}
times <- compare_times_simulation(N = N, m = m_rat, range = range, 
sigma = sigma, nsim = 1, samples = 100, m_nngp_fun = m_nngp_fun_sim, m_pca_fun = m_pca_fun_sim)
```
